Accelerated Inference for Partially Observed Stochastic Processes with Automatic Differentiation

### 2024-03-25 mop/draft.tex
Updates from Overleaf

\begin{aspt}[Bounded Measurement Model]
\label{assump:bounded-measurement}
There exist $\underbar{G}, \bar{G}$ such that $0<\underbar{G} \leq f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right) \leq \bar{G} < \infty$ and $\infty > G' \geq G'(\theta)>0$ such that $||\nabla_\theta \log f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right)||_\infty < G'(\theta) \leq G'$ for all $1 \leq n \leq N, x_n \in \gX$ and $\theta \in \Theta.$
\end{aspt}

\begin{aspt}[Locally Bounded Derivatives]
\label{assump:local-bounded-derivative}
Let $\mathcal{M}$ be an open subset of $\Omega$. There exists some function $G(\theta)$ and constants $G=\sup _{\theta \in \mathcal{N}} G(\theta)<\infty$, $H=\sup _{\theta \in \mathcal{N}} H(\theta)<\infty$ such that
    $$
    \|\nabla_\theta \ell(\theta, \phi, \omega, J)\|_\infty < NG'(\theta) \leq G(\theta) \leq G<\infty, \;\;
    \|\nabla^2_\theta \ell(\theta, \phi, \omega, J)\|_\infty <N^2G'(\theta)^2 \leq G(\theta)^2 \leq G^2<\infty
    $$
    for every $\phi \in \mathcal{M}, \theta \in \mathcal{N}_J(\phi)$, and almost every $\omega \in \Omega$.
\end{aspt}

@@

@@

@@

Therefore,
\begin{align*}
    \text{Var}(\nabla_\theta \hat\ell(\theta) &= \frac{1}{J^2}\text{Var}\left(\sum_{n=1}^N\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
    &= \frac{1}{J^2}\sum_{n=1}^N\text{Var}\left(\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
    &+ 2\sum_{m<n}\text{Cov}\left(\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_m|X_{m}}(y_m^*|x_{m,j}^{F, \theta}; \theta)\right), \sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
    &= \sum_{n=1}^N \text{Var}(\nabla_\theta\hat\ell_n(\theta)) + 2\sum_{m<n} \text{Cov}(\nabla_\theta\hat\ell_m(\theta),\nabla_\theta\hat\ell_n(\theta)).
\end{align*}

Here, we use Assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} that ensure strong mixing. We know from Theorem 3 of \cite{karjalainen2023} that when $\textbf{M}_{n,n+k}$ is the $k$-step Markov operator from timestep $n$ and $\beta_{\text{TV}}(M) = \sup _{x, y \in E}\|M(x, \cdot)-M(y, \cdot)\|_{\mathrm{TV}}=\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu M-\nu M\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}}$ is the Dobrushin contraction coefficient of a Markov operator, 
$$\beta_{\text{TV}}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},$$
i.e. the mixing time of the particle filter is $O(\log(J))$.

The particle filter itself is therefore strongly mixing, with $\alpha$-mixing coefficients $\alpha(k) \leq (1-\epsilon)^{\floor{k/(c\log(J))}}$. Therefore, functions of particles are strongly mixing as well, with $\alpha$-mixing coefficients bounded by the original (to see this, observe that the $\sigma$-algebra of the functionals is contained within the original $\sigma$-algebra). Therefore, by Davydov's inequality and noting that $\nabla_\theta\hat\ell_n(\theta)\leq G'(\theta)$ by Assumption \ref{assump:bounded-measurement},

$$\text{Cov}(\nabla_\theta\hat\ell_m(\theta), \nabla_\theta\hat\ell_n(\theta)) \leq G'(\theta)^2(1-\epsilon)^{\floor{\frac{|n-m|}{2c\log(J)}}}.$$

We can then conclude with Popoviciu's inequality that 
\begin{align*}
    \text{Var}(\nabla_\theta \hat\ell(\theta)) &= \sum_{n=1}^N \text{Var}(\nabla_\theta\hat\ell_n(\theta)) + 2\sum_{m<n} \text{Cov}(\nabla_\theta\hat\ell_m(\theta),\nabla_\theta\hat\ell_n(\theta))\\
    &\leq \frac{NG'(\theta)^2}{4J} + 2\sum_{n=1}^{N-1} (N-n) G'(\theta)^2(1-\epsilon)^{\floor{\frac{n}{2c\log(J)}}}\frac{G'(\theta)^2}{4J} \\
    &\leq \frac{G'(\theta)^2}{4J} \left(N + 2\sum_{n=1}^{N-1} (N-n) G'(\theta)^2(1-\epsilon)^{\floor{\frac{n}{2c\log(J)}}}\right) \\
    &\lesssim \frac{NG'(\theta)^2}{J}.
\end{align*}

\paragraph{Remark:} Note that the factor of $N$ that pops up here is due to the use of the unnormalized gradient. If one divides the gradient estimate by $N$ before usage, the variance does not depend on the horizon.

@@

Next, we analyze the case for $\alpha \in (0,1)$.

Notice that
\begin{align*}
    w_{n,j}^{F,\theta}
    &= \left(w_{n-1,k_j}^{F, \theta}\right)^\alpha\frac{g_{n,k_j}^\theta}{g_{n,k_j}^\phi} \\
    &= \left(\left(w_{n-2,j}^{A,F, \theta}\right)^\alpha\frac{g_{n-1,j}^{A,F,\theta}}{g_{n-1,j}^{A,F,\phi}}\right)^\alpha\frac{g_{n,j}^{A,F,\theta}}{g_{n,j}^{A,F,\phi}} \\
    &= \prod_{i=1}^n \left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}.
\end{align*}

Then, we have that
\begin{align*}
    \nabla_\theta\hat\ell(\theta) 
    &= \sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^Jw_{n,j}^{F,\theta}}{\sum_{j=1}^J\left(w_{n-1,j}^{F, \theta}\right)^\alpha}\right)= \sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}{\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}\right).
\end{align*}

Consider the modified estimator, where we truncate weights beyond $k$ timesteps:
\begin{align*}
    \hat{s}_k(\theta) 
    &:= \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}\right).
\end{align*}

Write $$w_{n,j}^{F,\theta,k} := \prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}$$ for the $k$-truncated weights.

By Assumption \ref{assump:bounded-measurement} we have that

\begin{align*}
   w_{n,j}^{F,\theta} 
    &\leq w_{n,j}^{F,\theta,k} \prod_{i=1}^{n-k}\left(\frac{\bar{G}}{\underbar{G}}\right)^{\alpha(n-i)}\\
    &\leq w_{n,j}^{F,\theta,k} \left(\frac{\bar{G}}{\underbar{G}}\right)^{\alpha \sum_{i=1}^{n-k}(n-i)} \\
    &\leq w_{n,j}^{F,\theta,k} \left(\frac{\bar{G}}{\underbar{G}}\right)^{\alpha\frac{n(n-1)-k(k-1)}{2}}
\end{align*}

We can then bound $||\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta)||$:
\begin{align*}
    &\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta) \\
    &= \sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}{\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}\right) - \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}\right) \\
    &= \sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}\right)- \log\left(\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}\right)\\
    &\qquad\qquad -\log\left(\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}\right) + \log\left(\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}\right)\Bigg) \\
    &= \sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,k}\right)
    -\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta}\right) + \log\left(\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}\right)\Bigg)
\end{align*}

We first bound $||\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta)||$:
\begin{align*}
    &\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta) \\
    &= \sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}{\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}\right) - \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}\right) \\
    &= \sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}\right)- \log\left(\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}\right)\\
    &\qquad\qquad -\log\left(\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}\right) + \log\left(\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}\right)\Bigg) \\
    &=  \sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}{\sum_{j=1}^J\prod_{i=n-k}^{n}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}\right) + \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}{\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}\right) 
\end{align*}

>>>>

\begin{aspt}[Bounded Measurement Model]
\label{assump:bounded-measurement}
There exist $\underbar{G}, \bar{G}$ such that $0<\underbar{G} \leq f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right) \leq \bar{G} < \infty$ and $\infty > G' \geq G'(\theta)>0$ such that $||\nabla_\theta \log f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right)||_\infty <\leq G'(\theta) \leq G'$ for all $1 \leq n \leq N, x_n \in \gX$ and $\theta \in \Theta.$
\end{aspt}

\begin{aspt}[Locally Bounded Derivatives]
\label{assump:local-bounded-derivative}
Let $\mathcal{M}$ be an open subset of $\Omega$. There exists some function $G(\theta)$ and constants $G=\sup _{\theta \in \mathcal{N}} G(\theta)<\infty$, $H=\sup _{\theta \in \mathcal{N}} H(\theta)<\infty$ such that
    $$
    \|\nabla_\theta \ell(\theta, \phi, \omega, J)\|_\infty \leq G(\theta) \leq G<\infty, \;\;
    \|\nabla^2_\theta \ell(\theta, \phi, \omega, J)\|_\infty \leq G(\theta)^2 \leq G^2<\infty
    $$
    for every $\phi \in \mathcal{M}, \theta \in \mathcal{N}_J(\phi)$, and almost every $\omega \in \Omega$.
\end{aspt}

@@

\paragraph{Acknowledgements.} We thank Nicholas Chopin for helpful comments on the proof of a strong law of large numbers for a suitably reweighted particle filter.

\kevin{Reach: PNAS. Else try for JRSSB, try to get Chopin for Annals AE. Would need to change bracketing conventions for Biometrika. If submitting to PNAS, JASA, Biometrika, or JRSSB, highlight the different statistical gradient estimator and finite sample performance, rewrite introduction, maybe create a way to use the gradients for inference rather than using profiling. Try to rewrite it as a breakthrough in statistics. Safety: JMLR}

@@

\subsection{Analysis of MOP-$0$ Variance}

@@

Therefore,
\begin{align*}
    \text{Var}(\nabla_\theta \hat\ell(\theta)) &= \frac{1}{J^2}\text{Var}\left(\sum_{n=1}^N\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
    &= \frac{1}{J^2}\sum_{n=1}^N\text{Var}\left(\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
    &+ 2\sum_{m<n}\text{Cov}\left(\frac{1}{J}\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_m|X_{m}}(y_m^*|x_{m,j}^{F, \theta}; \theta)\right), \frac{1}{J}\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
    &= \sum_{n=1}^N \text{Var}(\nabla_\theta\hat\ell_n(\theta)) + 2\sum_{m<n} \text{Cov}(\nabla_\theta\hat\ell_m(\theta),\nabla_\theta\hat\ell_n(\theta)).
\end{align*}

Here, we use Assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} that ensure strong mixing. We know from Theorem 3 of \cite{karjalainen2023} that when $\textbf{M}_{n,n+k}$ is the $k$-step Markov operator from timestep $n$ and $\beta_{\text{TV}}(M) = \sup _{x, y \in E}\|M(x, \cdot)-M(y, \cdot)\|_{\mathrm{TV}}=\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu M-\nu M\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}}$ is the Dobrushin contraction coefficient of a Markov operator, 
$$\beta_{\text{TV}}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},$$
i.e. the mixing time of the particle filter is $O(\log(J))$, where $\epsilon$ and $c$ depend on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ in \ref{assump:bounded-process} and \ref{assump:bounded-measurement}.

\kevin{How does the dobrushin contraction coefficient relate to the alpha-mixing coefficients? If you have a bound on the mixing time $b$, then the alpha mixing coefficients should just be $\alpha(t) \leq e^{-t/b}$.}

By Lemma \ref{lem:dobrushin-implies-alpha-mixing}, the particle filter itself is strongly mixing, with $\alpha$-mixing coefficients $\alpha(k) \leq (1-\epsilon)^{\floor{k/(c\log(J))}}$. Therefore, functions of particles are strongly mixing as well, with $\alpha$-mixing coefficients bounded by the original (to see this, observe that the $\sigma$-algebra of the functionals is contained within the original $\sigma$-algebra). Therefore, by Davydov's inequality, noting that $\nabla_\theta\hat\ell_n(\theta)\leq G'(\theta)$ by Assumption \ref{assump:bounded-measurement}, and WLOG assuming $\E[(\nabla_\theta\hat\ell_m(\theta))^4]^{1/4}\leq\E[(\nabla_\theta\hat\ell_n(\theta))^4]^{1/4}$, we see that
\begin{align*}
    \text{Cov}(\nabla_\theta\hat\ell_m(\theta), \nabla_\theta\hat\ell_n(\theta)) 
    &\leq \alpha(n-m)^{1/2}\E[(\nabla_\theta\hat\ell_m(\theta))^4]^{1/4}\E[(\nabla_\theta\hat\ell_n(\theta))^4]^{1/4}\\
    &\leq \alpha(n-m)^{1/2}\E[(\nabla_\theta\hat\ell_n(\theta))^4]^{1/2}.
\end{align*}

To bound this, we use the fact that $$\E[(\nabla_\theta\hat\ell_n(\theta))^4] = \E[(\nabla_\theta\hat\ell_n(\theta)-\E[\nabla_\theta\hat\ell_n(\theta)])^4]+\E[(\nabla_\theta\hat\ell_n(\theta))^2]^2$$
alongside Theorem 2 of \cite{delMoral03}, which shows that
$$\E[(\nabla_\theta\hat\ell_n(\theta)-\E[\nabla_\theta\hat\ell_n(\theta)])^4] \lesssim \frac{c_p^4G'(\theta)^4}{J^2}, \;\; \E[(\nabla_\theta\hat\ell_n(\theta)-\E[\nabla_\theta\hat\ell_n(\theta)])^2] \lesssim \frac{c_p^2G'(\theta)^2}{J},$$
where $c_p$ is a term depending only on the dimension of the state space $p$ (but where the dependence may be exponential), to find that 
$$\E[(\nabla_\theta\hat\ell_n(\theta))^4]^{1/2} \lesssim  \sqrt{\frac{c_p^4G'(\theta)^4}{J^2}+\left(\frac{c_p^2G'(\theta)^2}{J}\right)^2} = \frac{c_p^2G'(\theta)^2}{J},$$
and conclude that 
$$\text{Cov}(\nabla_\theta\hat\ell_m(\theta), \nabla_\theta\hat\ell_n(\theta)) \leq (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c\log(J)}}}\frac{c_p^2G'(\theta)^2}{J}.$$

Putting it all together, we see that
\begin{align*}
    \text{Var}(\nabla_\theta \hat\ell(\theta)) &= \sum_{n=1}^N \text{Var}(\nabla_\theta\hat\ell_n(\theta)) + 2\sum_{m<n} \text{Cov}(\nabla_\theta\hat\ell_m(\theta),\nabla_\theta\hat\ell_n(\theta))\\
    &\leq \frac{Nc_p^2G'(\theta)^2}{J} + 2\sum_{n=1}^{N-1} (N-n)(1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c\log(J)}}}\frac{c_p^2G'(\theta)^2}{J} \\
    &\leq \frac{c_p^2G'(\theta)^2}{J} \left(N + 2\sum_{n=1}^{N-1} (N-n) (1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c\log(J)}}}\right) \\
    &\lesssim \frac{c_p^2G'(\theta)^2N}{J}.
\end{align*}

\paragraph{Remark:} Note that the factor of $N$ that pops up here is due to the use of the unnormalized gradient. If one divides the gradient estimate by $\sqrt{N}$ before usage, the variance does not depend on the horizon. If one divides by $N$, the error is $O(1/\sqrt{NJ})$!

@@

\subsection{Analysis of MOP-$\alpha$ Variance}

Next, we analyze the case for $\alpha \in (0,1)$. The proof strategy is as follows. Instead of analyzing the variance of the MOP-$\alpha$ gradient estimate proper, we will analyze the variance of a modified estimator where we truncate the weights at $k$ timesteps, so this estimator only looks at most $k$ timesteps back while accumulating the importance weights. This enables us to establish strong mixing for the truncated estimator, in order to get a bound on the variance that is $O(NJ^{-1})$. We emphasize that this estimator is only a theoretical construct -- we do not actually use this estimator.

Coupled with a bound ensuring that the $k$-truncated estimator provides an estimate that is close to that of MOP-$\alpha$ proper, we get a bound on the variance of MOP-$\alpha$ that comprises of the variance of the $k$-truncated estimator plus the error, for any $k \leq N$. It then holds that the final variance bound on MOP-$\alpha$ is given by the minimum over all $k$, and is never larger than $O(1+NJ^{-1})$.

Using the log-derivative trick and that $w_{n,j}^{F,\theta} = w_{n,j}^{F,\theta,k} = 1$ when $\theta=\phi$,
\begin{align*}
    &\left\lVert\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta}\right)-\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,k}\right)\right\rVert_{\infty}\\
    &= \left\lVert\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta}}{{\sum_{j=1}^J w_{n,j}^{F,\theta}}}-\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta,k}}{{\sum_{j=1}^J w_{n,j}^{F,\theta,k}}}\right\rVert_{\infty} \\
    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta}-\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta,k}\right\rVert_{\infty} \\
    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta}}{w_{n,j}^{F,\theta}}-\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta,k}}{w_{n,j}^{F,\theta,k}}\right\rVert_{\infty} \\
    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta}\right)-\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta,k}\right)\right\rVert_{\infty} \\
    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\log\left(w_{n,j}^{F,\theta}\right)-\log\left(w_{n,j}^{F,\theta,k}\right)\right)\right\rVert_{\infty}\\
    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=1}^n\alpha^{(n-i)}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) - \sum_{i=n-k}^n{\alpha^{(n-i)}}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)\right\rVert_{\infty} \\
    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=1}^{n-k}\alpha^{(n-i)}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) \right)\right\rVert_{\infty}\\
    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=1}^{n-k}\alpha^{(n-i)}\left\lVert\nabla_\theta\log\left(g_{i,j}^{A,F,\theta} \right)\right\rVert_{\infty}\\
    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=1}^{n-k}\alpha^{(n-i)}G'(\theta)\\
    &\leq G'(\theta)\frac{\alpha^k-\alpha^n}{1-\alpha},
\end{align*}
where the second-last line follows from Assumption \ref{assump:bounded-measurement}.

We can then bound $||\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta)||$:
\begin{align*}
    &\left\lVert\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta) \right\rVert_\infty\\
    &= \left\lVert\sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) - \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) \right\rVert_\infty\\
    &= \Bigg\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right)- \log\left(\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right)\\
    &\qquad\qquad -\log\left(\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right) + \log\left(\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right)\Bigg)\Bigg\rVert_\infty \\
    &= \left\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,k}\right)
    -\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,k}\right)\Bigg)\right\rVert_\infty \\
    &\leq \sum_{n=1}^N \left\lVert\nabla_\theta \left(\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,k}\right)\right)\right\lVert_{\infty}
    +\sum_{n=1}^N \left\lVert\nabla_\theta \left(\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,k}\right)\right)\right\rVert_\infty,
\end{align*}
and each of these two terms is bounded by $\frac{\alpha^k}{1-\alpha}NG'(\theta)$. So we now know that
$$\left\lVert\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta) \right\rVert_\infty \leq  \frac{2\alpha^k}{1-\alpha}NG'(\theta),$$
which is our desired error bound.

Now, we bound the variance of $\hat s_k(\theta)$. 
Recall that we defined
$$\hat s_k(\theta) = \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) = \sum_{n=1}^N \nabla_\theta \left(\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta, k}\right)-\log\left(\sum_{j=1}^J w_{n-1,j}^{A, F,\theta, k}\right)\right).$$

We can then decompose this expression with the log-derivative trick while noting that $w_{n,j}^{F,\theta, k} = 1$ whenever $\theta=\phi$, to see that
\begin{align*}
    \hat s_k(\theta) 
    &= \sum_{n=1}^N \nabla_\theta \left(\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta, k}\right)-\log\left(\sum_{j=1}^J w_{n-1,j}^{A, F,\theta, k}\right)\right) \\
    &= \sum_{n=1}^N \left(\frac{\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta, k}}{\sum_{j=1}^J w_{n,j}^{F,\theta, k}}-\frac{\sum_{j=1}^J \nabla_\theta w_{n-1,j}^{A,F,\theta, k}}{\sum_{j=1}^J w_{n-1,j}^{A,F,\theta, k}}\right) \\
    &= \sum_{n=1}^N \left(\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta, k}-\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n-1,j}^{A,F,\theta, k}\right)\\
    &= \frac{1}{J}\sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \left(w_{n,j}^{F,\theta, k}- w_{n-1,j}^{A,F,\theta, k}\right).
\end{align*}

It now follows that we need to bound the variance at a single timestep, namely
$$\text{Var}\left(\frac{1}{J}\sum_{j=1}^J\nabla_\theta \left(w_{n,j}^{F,\theta, k}- w_{n-1,j}^{A,F,\theta, k}\right)\right).$$

We use the log-derivative trick yet again to find, noting that $w_{n,j}^{F,\theta, k}=1$ when $\theta=\phi$, that
\begin{align*}
    \nabla_\theta w_{n,j}^{F,\theta, k} = \frac{\nabla_\theta w_{n,j}^{F,\theta, k}}{w_{n,j}^{F,\theta, k}} = \nabla_\theta \log(w_{n,j}^{F,\theta, k}) = \sum_{i=n-k}^n \alpha^{n-i} \nabla_\theta \log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) = \sum_{i=n-k}^n \alpha^{n-i} \nabla_\theta \log\left(g_{i,j}^{A,F,\theta}\right).
\end{align*}
